\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-30 paper.qmd
Add preprocessing steps with Festem

### Human peripheral blood mononclear cells (PBMCs) data

@@

- Numb bins along the x-axis = 23, shape parameter = 0.8772751

@@

error_df <- read_csv("data/pbmc/pbmc_3k_festem/error_pbmc_umap.csv")

error_df |>
  knitr::kable() 
```

@@

shape_value_pbmc <- calculate_effective_shape_value(.data = UMAP_pbmc,
                                                      x = UMAP1, y = UMAP2) ## 0.8772751

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
##| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + 
plot_layout(ncol=2)

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 15, min_dist = 0.99, metric = euclidean) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP2
##| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + 
plot_layout(ncol=2)

@@

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP3
##| out-width: 100%
#| fig-pos: h

plot_list2_pbmc + trimesh_removed_pbmc_umap + 
plot_layout(ncol=2)

@@

>>>>

### Single-cell RNA-seq data of human

In the field of single-cell studies, a common analysis task involves clustering to identify groups of cells with similar expression profiles. Analysts often turn to NLDR techniques to verify and identify these clusters and explore developmental trajectories. In clustering workflows, the main objective is to verify the existence of clusters and subsequently identify them as specific cell types by examining the expression of "known" marker genes. Marker genes are specific genes or transcripts that are uniquely or highly expressed in particular cell types. Marker genes are identified through differential gene expression (DEG) analysis, comparing the expression levels of genes across different cell types. In this context, a "faithful" embedding should ideally preserve the topology of the data, ensuring that cells corresponding to the same cell type are situated close to the high-dimensional space.

### Processing steps

To begin our analysis, we installed the Peripheral Blood Mononuclear Cells (pbmc) data set obtained from 10x Genomics using the `SeuratData` R package [@Rahul2019], which facilitates the distribution of data sets in the form of Seurat objects [@Yuhan2021]. This data set contains 13,714 features (genes) across 2,700 samples (cells) within a single assay. The active assay is RNA, with 13,714 features representing different gene expressions. Then, the cells that have unique feature counts over 2,500 or less than 200 and cells that have > 5\% mitochondrial counts are filtered. After removing unwanted cells from the dataset, the next step is to normalize the data. By default, we employ a global-scaling normalization method “LogNormalize” that normalizes the feature expression measurements for each cell by the total expression, multiplies this by a scale factor (10,000 by default), and log-transforms the result.

Next, top 1000 genes are selected by Festem [@Chen2023] DEG method. Then, scale the data. Then, perform PCA on the scaled data. Louvain algorithm is used to cluster the single cells based on the genes selected by Festem. Using the Festem-selected genes and 15 PCs, identified 10 clusters in the PBMC3k data and annotated them based on the expression of canonical markers. The 10 clusters included immune cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and Dendritic Cells (DC). In addition to these common cell types, Festem identified a fine cell type, CD27− CD4+ memory T cells, which were often missed by other methods.

- Human peripheral blood mononclear cells (PBMCs) data

- Intro to the data set

Used Louvain algorithm to cluster the single cells based on the genes selected by Festem method

- Explain what "Evaluate Festem on DEG detection" means, in terms of this data

-

- How did they "identified 10 clusters"? What were the variables used, PCs or full data or 2D representation, spell this out.

- Used top 1000 genes selected by Festem to run PCA, top 15 PCs (original paper mentioned)

- The "10 clusters included immune cells such as naive CD4 T ..." does this mean that every cluster had all of these, or one cluster was mostly "CD4 T cells", another mostly "memory CD4 T", ...?

- How is your use of this data different from the original application.

- Original paper used this data set to compare different DEG methods, but we need to look how the performance of UMAP with the author's selected parameter choice.

- Why 15 PCs? Is that from the original paper? YES

@@

- Num bins along the x-axis = 23, shape parameter = 0.8772751

@@

- The third parameter is metric. This parameter plays a crucial role in determining how distances are computed within the ambient space of the input data (The term "ambient space" refers to the original or initial space in which the data exists before any transformation or dimensionality reduction).

# error_df <- read_csv("data/pbmc/pbmc_3k_festem/error_pbmc_umap.csv")
# 
# error_df |>
#   knitr::kable() 
```

| n_neighbors | min_dist | metric     | error                 |
|-------------|----------|------------|-----------------------|
| **30**      | **0.3**  | **cosine** | **38634.66719286174** |
| 5           | 0.99     | euclidean  | 35938.98594054407     |
| 5           | 0.99     | cosine     | 35597.01981261912     |
| 15          | 0.99     | euclidean  | 35534.75778935427     |
| 30          | 0.99     | euclidean  | 35664.68763936665     |
| 1022        | 0.99     | cosine     | 37022.0125249776      |
| 1022        | 0.3      | cosine     | 39746.347504639125    |
| 800         | 0.3      | euclidean  | 38385.25485420897     |
| 100         | 0.99     | euclidean  | 36528.57424747017     |
| 230         | 0.99     | cosine     | 37864.26802578186     |
| 354         | 0.99     | euclidean  | 37530.86598361975     |

@@

@@

df_bin_train <- df_bin_pbmc
names(df_bin_train)[-1] <- paste0("avg_", names(df_bin_train)[-1])

error_df <- df_all_pbmc |>
  dplyr::left_join(df_bin_train, by = c("hb_id" = "hb_id")) ## Map high-D averaged/weighted mean coordinates

# prediction_df_join <- prediction_df_join |>
#   dplyr::left_join(data, by = c("ID" = "ID")) ## Map high-D data

for (i in 1:(NCOL(df_bin_train) - 1)) {

  error_df[ , paste0("abs_residual_", "PC_", i)] <- abs(error_df[ , paste0("PC_", i)] - error_df[ , paste0("avg_", "PC_", i)])

error_df <- error_df |>
  dplyr::mutate(total = rowSums(dplyr::pick(tidyselect::starts_with(paste0("abs_residual_", "PC_")))))

# library(ggbeeswarm)
# error_df$group <- "1"
# ggplot(error_df, aes(x = group, y = total)) +
#   geom_quasirandom()+
#   ylim(0, max(unlist(error_df$total))+ 0.5) + coord_flip()

### The minimum error is 0 and the maximum is 42.17439
### There is lot of points with error 0,

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 2 or less",
                        if_else(total <= 10, "error 2-10",
                                if_else(total <= 15, "error 10-15",
                                        if_else(total <= 20, "error 15-20",
                                                if_else(total <= 25, "error 20-25",
                                                        if_else(total <= 30, "error 25-30",
                                                                if_else(total <= 35, "error 30-35", "error greter than 35"))))))))

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("error 2 or less", "error 2-10", "error 10-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error greter than 35"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) #+
  #geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  #annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 30, min_dist = 0.3) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<https://youtu.be/CNXjWZMyh_M>)"
#| label: fig-nldervisPBMCUMAP
#| out-width: 100%
#| fig-pos: h

# plot_list2_pbmc + 
trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')

@@

df_bin_train <- df_bin_pbmc
names(df_bin_train)[-1] <- paste0("avg_", names(df_bin_train)[-1])

error_df <- df_all_pbmc |>
  dplyr::left_join(df_bin_train, by = c("hb_id" = "hb_id")) ## Map high-D averaged/weighted mean coordinates

# prediction_df_join <- prediction_df_join |>
#   dplyr::left_join(data, by = c("ID" = "ID")) ## Map high-D data

for (i in 1:(NCOL(df_bin_train) - 1)) {

  error_df[ , paste0("abs_residual_", "PC_", i)] <- abs(error_df[ , paste0("PC_", i)] - error_df[ , paste0("avg_", "PC_", i)])

error_df <- error_df |>
  dplyr::mutate(total = rowSums(dplyr::pick(tidyselect::starts_with(paste0("abs_residual_", "PC_")))))

# library(ggbeeswarm)
# error_df$group <- "1"
# ggplot(error_df, aes(x = group, y = total)) +
#   geom_quasirandom()+
#   ylim(0, max(unlist(error_df$total))+ 0.5) + coord_flip()

### The minimum error is 0 and the maximum is 42.17439
### There is lot of points with error 0,

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 2 or less",
                        if_else(total <= 10, "error 2-10",
                                if_else(total <= 15, "error 10-15",
                                        if_else(total <= 20, "error 15-20",
                                                if_else(total <= 25, "error 20-25",
                                                        if_else(total <= 30, "error 25-30",
                                                                if_else(total <= 35, "error 30-35", "error greter than 35"))))))))

plot_list2_pbmc_error <- error_df |>
    mutate(type = factor(type , levels = c("error 2 or less", "error 2-10", "error 10-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error greter than 35"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 15, min_dist = 0.99, metric = euclidean) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP2
#| out-width: 100%
#| fig-pos: h

#plot_list2_pbmc + 
trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')

@@

df_bin_train <- df_bin_pbmc
names(df_bin_train)[-1] <- paste0("avg_", names(df_bin_train)[-1])

error_df <- df_all_pbmc |>
  dplyr::left_join(df_bin_train, by = c("hb_id" = "hb_id")) ## Map high-D averaged/weighted mean coordinates

# prediction_df_join <- prediction_df_join |>
#   dplyr::left_join(data, by = c("ID" = "ID")) ## Map high-D data

for (i in 1:(NCOL(df_bin_train) - 1)) {

  error_df[ , paste0("abs_residual_", "PC_", i)] <- abs(error_df[ , paste0("PC_", i)] - error_df[ , paste0("avg_", "PC_", i)])

error_df <- error_df |>
  dplyr::mutate(total = rowSums(dplyr::pick(tidyselect::starts_with(paste0("abs_residual_", "PC_")))))

# library(ggbeeswarm)
# error_df$group <- "1"
# ggplot(error_df, aes(x = group, y = total)) +
#   geom_quasirandom()+
#   ylim(0, max(unlist(error_df$total))+ 0.5) + coord_flip()

### The minimum error is 0 and the maximum is 42.17439
### There is lot of points with error 0,

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 2 or less",
                        if_else(total <= 10, "error 2-10",
                                if_else(total <= 15, "error 10-15",
                                        if_else(total <= 20, "error 15-20",
                                                if_else(total <= 25, "error 20-25",
                                                        if_else(total <= 30, "error 25-30",
                                                                if_else(total <= 35, "error 30-35", "error greter than 35"))))))))

plot_list2_pbmc_error <- error_df |>
    mutate(type = factor(type , levels = c("error 2 or less", "error 2-10", "error 10-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error greter than 35"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d"))

```{r}
#| echo: false
#| fig-cap: "(a) 2D layout from UMAP (n_neighbors = 5, min_dist = 0.99, metric = cosine) applied for the PBMC3k dataset. Is this a best representation of the original data?, (b)  Model in the 2D space with UMAP (<>)"
#| label: fig-nldervisPBMCUMAP3
#| out-width: 100%
#| fig-pos: h

#plot_list2_pbmc + 
trimesh_removed_pbmc_umap + plot_list2_pbmc_error +
plot_layout(guides='collect', ncol=3) &
  theme(legend.position='bottom')

@@

- Num bins along the x-axis = 43, shape parameter = 0.4466309

### 2024-01-30 paper.tex
Add preprocessing steps with Festem

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\hypertarget{human-peripheral-blood-mononclear-cells-pbmcs-data}{%
\subsection{Human peripheral blood mononclear cells (PBMCs)
data}\label{human-peripheral-blood-mononclear-cells-pbmcs-data}}

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  Numb bins along the x-axis = 23, shape parameter = 0.8772751
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\item
  The most important parameter is n\_neighbors - the number of
  approximate nearest neighbors used to construct the initial
  high-dimensional graph. It effectively controls how UMAP balances
  local versus global structure - low values will push UMAP to focus
  more on local structure by constraining the number of neighboring
  points considered when analyzing the data in high dimensions, while
  high values will push UMAP towards representing the big-picture
  structure while losing fine detail.
\item
  The second parameter we'll investigate is min\_dist, or the minimum
  distance between points in low-dimensional space. This parameter
  controls how tightly UMAP clumps points together, with low values
  leading to more tightly packed embeddings. Larger values of min\_dist
  will make UMAP pack points together more loosely, focusing instead on
  the preservation of the broad topological structure.
\end{itemize}

\begin{tabular}{r|r|l|r}
\hline
30 & 0.3 & cosine & 38634.66719286174\\
\hline
5 & 0.99 & euclidean & 35938.99\\
\hline
5 & 0.99 & cosine & 35597.02\\
\hline
15 & 0.99 & euclidean & 35534.76\\
\hline
30 & 0.99 & euclidean & 35664.69\\
\hline
1022 & 0.99 & cosine & 37022.01\\
\hline
1022 & 0.30 & cosine & 39746.35\\
\hline
800 & 0.30 & euclidean & 38385.25\\
\hline
\end{tabular}

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP2-1.pdf}

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-nldervisPBMCUMAP3-1.pdf}

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters, but the clusters are
  really close
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
  Also, the two clusters are really near to each other and can't see any
  separation. Also continuity of the structure. Long edges are presence
  also.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\end{itemize}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

\hypertarget{single-cell-rna-seq-data-of-human}{%
\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}}

In the field of single-cell studies, a common analysis task involves
clustering to identify groups of cells with similar expression profiles.
Analysts often turn to NLDR techniques to verify and identify these
clusters and explore developmental trajectories. In clustering
workflows, the main objective is to verify the existence of clusters and
subsequently identify them as specific cell types by examining the
expression of ``known'' marker genes. Marker genes are specific genes or
transcripts that are uniquely or highly expressed in particular cell
types. Marker genes are identified through differential gene expression
(DEG) analysis, comparing the expression levels of genes across
different cell types. In this context, a ``faithful'' embedding should
ideally preserve the topology of the data, ensuring that cells
corresponding to the same cell type are situated close to the
high-dimensional space.

\hypertarget{processing-steps}{%
\subsection{Processing steps}\label{processing-steps}}

To begin our analysis, we installed the Peripheral Blood Mononuclear
Cells (pbmc) data set obtained from 10x Genomics using the
\texttt{SeuratData} R package \citep{Rahul2019}, which facilitates the
distribution of data sets in the form of Seurat objects
\citep{Yuhan2021}. This data set contains 13,714 features (genes) across
2,700 samples (cells) within a single assay. The active assay is RNA,
with 13,714 features representing different gene expressions. Then, the
cells that have unique feature counts over 2,500 or less than 200 and
cells that have \textgreater{} 5\% mitochondrial counts are filtered.
After removing unwanted cells from the dataset, the next step is to
normalize the data. By default, we employ a global-scaling normalization
method ``LogNormalize'' that normalizes the feature expression
measurements for each cell by the total expression, multiplies this by a
scale factor (10,000 by default), and log-transforms the result.

Next, top 1000 genes are selected by Festem \citep{Chen2023} DEG method.
Then, scale the data. Then, perform PCA on the scaled data. Louvain
algorithm is used to cluster the single cells based on the genes
selected by Festem. Using the Festem-selected genes and 15 PCs,
identified 10 clusters in the PBMC3k data and annotated them based on
the expression of canonical markers. The 10 clusters included immune
cells such as naive CD4 T cells, memory CD4 T cells, CD8 T cells, CD14
monocytes, FCGR3A monocytes, Natural Killer (NK) cells, B cells and
Dendritic Cells (DC). In addition to these common cell types, Festem
identified a fine cell type, CD27− CD4+ memory T cells, which were often
missed by other methods.

\begin{itemize}
\item
  Human peripheral blood mononclear cells (PBMCs) data
\item
  Intro to the data set
\end{itemize}

Used Louvain algorithm to cluster the single cells based on the genes
selected by Festem method

\begin{itemize}
\item
  Explain what ``Evaluate Festem on DEG detection'' means, in terms of
  this data
\item
\item
  How did they ``identified 10 clusters''? What were the variables used,
  PCs or full data or 2D representation, spell this out.
\item
  Used top 1000 genes selected by Festem to run PCA, top 15 PCs
  (original paper mentioned)
\item
  The ``10 clusters included immune cells such as naive CD4 T \ldots{}''
  does this mean that every cluster had all of these, or one cluster was
  mostly ``CD4 T cells'', another mostly ``memory CD4 T'', \ldots?
\item
  How is your use of this data different from the original application.
\item
  Original paper used this data set to compare different DEG methods,
  but we need to look how the performance of UMAP with the author's
  selected parameter choice.
\item
  Why 15 PCs? Is that from the original paper? YES
\end{itemize}

@@

\begin{itemize}
\item
  Intro to the data set (This data set contains 13,714 features across
  2,700 samples within a single assay. The active assay is RNA, with
  13,714 features representing different gene expressions.)
\item
  2622 cells
\item
  Evaluate Festem on DEG detection (Festem enabled identification of
  often-missed fine cell types)
\item
  Using the Festem-selected genes, identified 10 clusters in the PBMCsk
  data and annotated them based on the expression of canonical markers
\item
  The 10 clusters included immune cells such as naive CD4 T cells,
  memory CD4 T cells, CD8 T cells, CD14 monocytes, FCGR3A monocytes,
  Natural Killer (NK) cells, B cells and Dendritic Cells (DC).
\item
  In addition to these common cell types, Festem identified a fine cell
  type, CD27− CD4+ memory T cells, which were often missed by other
  methods.
\item
  The CD27. CD4+ memory T cells identified by Festem expressed common
  marker genes (IL7R and S100A4) of memory T cells, but did not express
  CD27. These cells also had downregulated expression of SELL, CCR7, MAL
  and LEF1, and upregulated expression of CCL5 (Fig. 4B) and thus were
  the CD27- CD4+ memory T cells in the literature (36). The CD27- CD4+
  memory T cells were known to be at a more differentiated state and
  have stronger antigen-recall responses than their CD27+ counterparts.
\item
  15 PCs
\item
  Num bins along the x-axis = 23, shape parameter = 0.8772751
\item
  learned from the model: There are 3 well-separated clusters in 2D, but
  if look at the model in high-D space, the three clusters are much
  closer. Also, there is some continuity within the clustering structure
  and it didn't capture well. Therefore UMAP with n:neighbour 30 is not
  a good representation for this data set.
\item
  The most important parameter is n\_neighbors - the number of
  approximate nearest neighbors used to construct the initial
  high-dimensional graph. It effectively controls how UMAP balances
  local versus global structure - low values will push UMAP to focus
  more on local structure by constraining the number of neighboring
  points considered when analyzing the data in high dimensions, while
  high values will push UMAP towards representing the big-picture
  structure while losing fine detail.
\item
  The second parameter we'll investigate is min\_dist, or the minimum
  distance between points in low-dimensional space. This parameter
  controls how tightly UMAP clumps points together, with low values
  leading to more tightly packed embeddings. Larger values of min\_dist
  will make UMAP pack points together more loosely, focusing instead on
  the preservation of the broad topological structure.
\item
  The third parameter is metric. This parameter plays a crucial role in
  determining how distances are computed within the ambient space of the
  input data (The term ``ambient space'' refers to the original or
  initial space in which the data exists before any transformation or
  dimensionality reduction).
\end{itemize}

\begin{longtable}[]{@{}llll@{}}
\toprule\noalign{}
n\_neighbors & min\_dist & metric & error \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
\textbf{30} & \textbf{0.3} & \textbf{cosine} &
\textbf{38634.66719286174} \\
5 & 0.99 & euclidean & 35938.98594054407 \\
5 & 0.99 & cosine & 35597.01981261912 \\
15 & 0.99 & euclidean & 35534.75778935427 \\
30 & 0.99 & euclidean & 35664.68763936665 \\
1022 & 0.99 & cosine & 37022.0125249776 \\
1022 & 0.3 & cosine & 39746.347504639125 \\
800 & 0.3 & euclidean & 38385.25485420897 \\
100 & 0.99 & euclidean & 36528.57424747017 \\
230 & 0.99 & cosine & 37864.26802578186 \\
354 & 0.99 & euclidean & 37530.86598361975 \\
\end{longtable}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP2-1.pdf}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-nldervisPBMCUMAP3-1.pdf}

@@

\begin{itemize}
\item
  In 2D, there are two well-separated clusters, but the clusters are
  really close
\item
  One cluster have a sparse data space within the cluster
\item
  There is a line in right corner of the structure that is not connected
  any other trimesh points. This line create a long edge in high-D
  space.
\item
  In high-D view, the clusters having some non-linearity patter, which
  is also not captured in tSNE. Therefore, this is not a good choice.
  Also, the two clusters are really near to each other and can't see any
  separation. Also continuity of the structure. Long edges are presence
  also.
\item
  The parameter is, in a sense, a guess about the number of close
  neighbors each point has.
\item
  Num bins along the x-axis = 43, shape parameter = 0.4466309
\end{itemize}
